---
title: "624_wk2_E_Haley"
author: "Ethan Haley"
date: "1/30/2022"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(fpp3)
```

**1) Use the help function to explore what the series `gafa_stock`, `PBS`, `vic_elec` and `pelt` represent.**

```{r}
#help("gafa_stock")
#help("PBS")
#help("vic_elec")
#help("pelt")
```

**a) Use autoplot() to plot some of the series in these data sets.**

```{r}
autoplot(gafa_stock, .vars = Close)
```

```{r}
autoplot(vic_elec, .vars = Demand)
```

```{r}
autoplot(pelt, .vars = Hare)
```

How do the 2 animals' pelt numbers align by year?

```{r}
pelt %>%    # using the code provided for Q.3
  pivot_longer(-Year) %>% 
  ggplot(aes(x = Year, y = value, colour = name)) + 
  geom_line() + 
  labs(y='pelts')
```


**b) What is the time interval of each series?**  


```{r}
interval(pelt)
```

`gafa_stock`: 1 day  

`PBS`:  1 month  

`vic_elec`: 30 minutes  

`pelt`: 1 year


**2) Use filter() to find what days corresponded to the peak closing price for each of the four stocks in `gafa_stock`**  

```{r}
highs = tibble()
for (stock in unique(gafa_stock$Symbol)) {
  high = gafa_stock %>%
    filter(Symbol==stock) %>%
    filter(Close==max(Close))
  highs = bind_rows(highs, high)
}
data.frame(highs)  # displays better than a tibble
```

**3) Download the file `tute1.csv` from the book website**  


```{r}
sales = read.csv("https://raw.githubusercontent.com/ebhtra/msds-624/main/tute1.csv")
head(sales)
```
**Convert the data to time series**  

```{r}
mytimeseries <- sales %>%
  mutate(Quarter = yearmonth(Quarter)) %>%
  as_tsibble(index = Quarter)
head(mytimeseries)
```
**Construct time series plots of each of the three series**  


```{r}
mytimeseries %>%
  pivot_longer(-Quarter) %>%
  ggplot(aes(x = Quarter, y = value, colour = name)) +
  geom_line() +
  facet_grid(name ~ ., scales = "free_y")
```

**6) Create time plots of the following four time series: `Bricks` from `aus_production`, `Lynx` from `pelt`, `Close` from `gafa_stock`, `Demand` from `vic_elec`.**

**Use ? (or help()) to find out about the data in each series.**
**For the last plot, modify the axis labels and title.**  

```{r}

```








